---
title: "Practice Lecture 16 MATH 342W Queens College"
author: "Professor Adam Kapelner"
---



# Log transformations

We will be examining the diamonds dataset. Let's take a moment to read about our data. In real life, you will take more than just a moment to learn about the data. And more than just a moment to clean the data and do sanity checks. As these steps are more of the grunt work of data science (and are learned on the fly), I won't really cover them formally.

```{r}
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
diamonds$cut =      factor(diamonds$cut, ordered = FALSE)      #convert to nominal
diamonds$color =    factor(diamonds$color, ordered = FALSE)    #convert to nominal
diamonds$clarity =  factor(diamonds$clarity, ordered = FALSE)  #convert to nominal
diamonds = diamonds[sample(1 : nrow(diamonds)), ] #shuffle for better visualizations
dim(diamonds)
```

That's a huge $n$. So, let's expect things to take a bit longer when processing.

A natural increasing relationship will likely be found between weight and price. Let's see it visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```


How good does a best guess linear relationship do?

```{r}
mod = lm(price ~ carat, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

What does the intercept say about extrapolation?

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green") + ylim(0, 50000)
```

Maybe the relationship between weight and price is not linear - but exponential? E.g. if the weight of a diamond doubles, maybe the price quadruples? Or linear increases in weight yield percentage increases in price. This seems plausible. 

Besides the plausibility of the functional form, there is an agnostic reason to employ log y as the predictive target. Let's first examine the univariate data!


```{r}
skimr::skim(diamonds$price)
```

Very large standard error and very long tail which can be seed more clearly here:

```{r}
ggplot(diamonds) + geom_histogram(aes(price), binwidth = 200)
```

Let's take a look at the distribution after logging:

```{r}
ggplot(diamonds) + geom_histogram(aes(x = log(price)), binwidth = 0.01)
```

Some strange artifacts appear. Why the gap? Why is it "cut" sharply at a maximum? These are questions to ask the one who collected the data. But let's get back to the log story...

Popular wisdom says logging this type of highly skewed-right distribution would possibly make the model "more linear in x". Put another way, it would be easier to "catch" (predict) the long tail since it won't be a long tail anymore after you log-transform. It would also prevent observations with large y's becoming "leverage points" i.e. points that unduly influence the model and thereby warp its ability to predict the average observation. In 343, you'll learn more reasons for why you should use log the response, but those reasons are relevant for inference so we won't discuss them here.

Let's give the model with ln(y) a whirl. Maybe we'll even learn something about diamonds. The way to create such a model is to simply fit an OLS model to log y. This is called a log-linear model. Since this is a pretty standard thing to do so R's formula notation has it built-in as follows:

```{r}
log_linear_mod = lm(log(price) ~ carat, diamonds)
b = coef(log_linear_mod)
b
```

Let's see what this looks like.

```{r}
ggplot(diamonds, aes(x = carat, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green") + ylim(6, 12)
```

It looks very bad if carat is large. That means our little theory about carats getting larger yielding multiples of price doesn't correspond to reality.

How did we do?

```{r}
summary(log_linear_mod)$r.squared
summary(log_linear_mod)$sigma
```

Look at that RMSE! That dropped like a rock! Is that real?

No. RMSE before is in the units of y. And now y is now in ln($). So this RMSE and the previous RMSE are *not* comparable.

The $R^2$ are *not* comparable either. Even though they're on a [0, 1] scale in both models, the SST's are different so you're measuring the proportion of a different variance.

Let's attempt to compare apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

What does this mean? It means this was a bad idea. Those residuals for large carats are insanely large. They're wrong on a log scale! Which means they're off by orders of magnitude. Working with logged y is dangerous business if you're wrong! Before you were off by a few thousand dollars; now you're off by millions. For example. Let's look at a large diamond:

```{r}
xstar = diamonds[diamonds$carat > 5, ][1, ]
xstar$price
predict(mod, xstar)
exp(predict(log_linear_mod, xstar))
```

That's a pretty bad residual!

How about log-log model? 

```{r}
log_log_linear_mod = lm(log(price) ~ log(carat), diamonds)
b = coef(log_log_linear_mod)
b
```

Let's see what it looks like:

```{r}
ggplot(diamonds, aes(x = log(carat), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Well look at that! That's a nice looking model. (Note that the slope coefficients in log-log models, i.e. b_2 here, are called "elasticity" in Econ 382 as it measures how the relative change in x affects the relative change in y).

How are our metrics?

```{r}
summary(log_log_linear_mod)$r.squared
summary(log_log_linear_mod)$sigma
```

Let's see apples-to-apples to the natural y model.

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

This is on-par with the vanilla OLS model, but still doesn't "beat it". There was no guarantee that we would be "beat it" even though we used procedures that are reasonable and popular!

Let's repeat this entire exercise using the length of the diamond. The length of the diamond feature is confusingly named "x" in the dataset. It is an "x" but it's also the diamond's "x"!!!

```{r}
ggplot(diamonds, aes(x = x, y = price)) + 
  geom_point()
```

Besides the non-linear relationship, what else do you see? Mistakes in the dataset! Can a real diamond have zero length?? Yes. This is the real world. There are mistakes all the time.

How many are we dealing with here?

```{r}
nrow(diamonds[diamonds$x == 0, ])
```
Let's kill it! That is let's *clean our data*!

```{r}
diamonds = diamonds[diamonds$x != 0, ]
```

How good does a best guess linear relationship do now?

```{r}
mod = lm(price ~ x, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = x, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Again we got some bad extrapolation going on which we can't fix using a purely linear modeling strategy.

Let's log-linearize it and see how we do.

```{r}
log_linear_mod = lm(log(price) ~ x, diamonds)
b = coef(log_linear_mod)
ggplot(diamonds, aes(x = x, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? Ensure it's apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

Still not better. Log-log?

```{r}
log_log_linear_mod = lm(log(price) ~ log(x), diamonds)
b = coef(log_log_linear_mod)
ggplot(diamonds, aes(x = log(x), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? 

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

We did it. We found a log transformation that seems to give higher predictive power than the vanilla linear model on the raw response and raw feature.

This brings up the whole idea of "model selection". We went hunting for models until we found one that's better.

Transforming y is a big decision as it changes the response metric! The rule of thumb is it is easier to model a response metric that has less extreme values (especially when using linear models) as the extreme values have a big impact on slope coefficients and can distort the best fit line due to the least squares minimization (hence the popularity of logging the response).

Let's see if we get anywhere with this using all the features in this model.

```{r}
lm_y = lm(price ~ ., diamonds)
lm_ln_y = lm(log(price) ~ ., diamonds)
summary(lm_y)$r.squared
summary(lm_y)$sigma

y_hat = exp(lm_ln_y$fitted.values)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
``` 

This is pretty convincing evidence that this transformation does a better job (at least in our linear modeling context).

Let's look at one prediction:

```{r}
predict(lm_y, diamonds[12345, ])
exp(predict(lm_ln_y, diamonds[12345, ]))
diamonds$price[12345]
```

Again, we should be careful when you use $g$ after logging, you will have to exponentiate the result (middle line above). 

Small point: this exponentiation is known to create bias because $E[Y]$ is different from $exp(E[ln(y)])$ (for those who took 368 - remember Jensen's inequality?) For the purposes of this class, this can be ignored since we are evaluating g on its own merits and we're doing so honestly.

If you like this stuff, there are a whole bunch of transformations out there that are even cooler than the natural log. 


# Linear Models with Feature Interactions

Let's go back to modeling price with weight. Let us add a third variable to this plot, color, a metric about the "yellowness" of the diamond. This is an ordinal categorical variable ranging from D (most clear i.e. best) to J (most yellow in this dataset i.e. worst).


```{r}
pacman::p_load(ggplot2)
base = ggplot(diamonds, aes(x = carat, y = price)) 
base +
  geom_point(aes(col = color)) + 
  scale_color_brewer(type = "div")
```

We can split the data on color to see it more clearly:

```{r}
base +
  geom_point() +
  facet_wrap(~ color, ncol = 3) + 
  aes(color = color) + scale_color_brewer(type = "div")
```


What do we see here? It looks like the slope of the price vs. carat linear model is slightly affected by color. For instance, the "D" color diamonds' price increases much faster as weight increases than the "E" color diamonds' price increases in weight, etc. Why do you think this is?

We can investigate two of these linear models below by fitting two submodels, one for D and one for J:

```{r}
mod_D = lm(price ~ carat, diamonds[diamonds$color == "D", ])
b_D = coef(mod_D)
mod_J = lm(price ~ carat, diamonds[diamonds$color == "J", ])
b_J = coef(mod_J)
b_D
b_J
```

Let's see it on the plot:

```{r}
base +
  geom_point(aes(col = color)) + scale_color_brewer(type = "div") +
  geom_abline(intercept = b_D[1], slope = b_D[2], col = "blue", lwd = 2) +
  geom_abline(intercept = b_J[1], slope = b_J[2], col = "red", lwd = 2)
```

This indicates a separate intercept and carat-slope for each color. How is this done? Interacting carat and slope. The formula notation has the `*` operator for this. It is a type of abstract multiplication after all!

```{r}
mod = lm(price ~ color, diamonds)
coef(mod)
mod = lm(price ~ carat * color, diamonds)
coef(mod)
```

The reference category is color D. This means every other color should start lower and have a lower slope. This is approximately what we see above.

How much of a better model is this than a straight linear model?

```{r}
mod_vanilla = lm(price ~ carat + color, diamonds)
summary(mod_vanilla)$r.squared
summary(mod_vanilla)$sigma
summary(mod_vanilla)$df[1]
summary(mod)$r.squared
summary(mod)$sigma
summary(mod)$df[1]
```

You can get more predictive accuracy out of this. 

Why did degrees of freedom increase? 

Is this gain real? Yes. With six more df's and $n = 54,000$ there is no chance this gain came from overfitting noise. Add 10,000 garbage features, that's a different story.

Let's take a look at carat with another variable, depth, a continuous predictor. High depth indicates diamonds are skinny and tall; low depth indicates diamonds are flat like a pancake.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = depth), lwd = 1, alpha = 0.5) + scale_colour_gradientn(colours = rainbow(5))
```

It seems people like flatter diamonds and are willing to pay more per carat. Let's see this in the regression:

```{r}
mod = lm(price ~ carat + depth, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * depth, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Also a real performance boost.

How about cut?


```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = cut), lwd = 0.5) + scale_color_brewer(type = "div")
```

Looks like ideal diamonds increase more steeply with size vs lower cut diamonds. Let's see this in action:

```{r}
mod = lm(price ~ carat, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Yes.

Can we include all these interactions?

```{r}
mod = lm(price ~ carat + color + depth + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * (color + depth + cut), diamonds)
summary(mod)$r.squared
summary(mod)$sigma
coef(mod)
```

A decent gain for all three.

What does the design matrix look like there? What is $p$?

```{r}
Xmm = model.matrix(price ~ carat * (color + depth + cut), diamonds)
head(Xmm)
```


Can we take a look at interactions of two categorical variables?


```{r}
plot1 = ggplot(diamonds, aes(x = cut, y = color)) +
  geom_jitter(aes(col = price), lwd = 0.5) + scale_colour_gradientn(colours = rainbow(5))
plot1
```
Let's see what the regressions say:


```{r}
mod = lm(price ~ color + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ color * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Not too much gain. Probably not worth it.


# More flexible H sets


```{r}
rm(list = ls())
pacman::p_load(ggplot2)
diamonds$cut =      factor(diamonds$cut, ordered = FALSE)
diamonds$color =    factor(diamonds$color, ordered = FALSE)
diamonds$clarity =  factor(diamonds$clarity, ordered = FALSE)
#drop some obviously nonsense observations
diamonds = diamonds[diamonds$carat <= 2 & diamonds$x != 0 & diamonds$y != 0 & diamonds$z != 0 & diamonds$depth != 0 & diamonds$table != 0,]
diamonds$ln_price = log(diamonds$price)
diamonds$ln_carat = log(diamonds$carat)
diamonds$ln_x = log(diamonds$x)
diamonds$ln_y = log(diamonds$y)
diamonds$ln_z = log(diamonds$z)
diamonds$ln_depth = log(diamonds$depth)
diamonds$ln_table = log(diamonds$table)
n = nrow(diamonds)
set.seed(1984)
diamonds = diamonds[sample(1 : n), ]
```

Note: I will now model price, not ln_price as ln_price yields residuals that are orders of magnitude. Fitting a good ln_price model will take more time.

```{r}
#All model formulas for reuse later
model_formulas = list(
  A = price ~ ln_carat,
  B = price ~ ln_carat * clarity,
  C = price ~ ln_carat * (clarity + cut + color),
  D = price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut + color),
  E = price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table + carat + x + y + z + depth + table) * (clarity + cut + color)
)
#Model A
mod = lm(model_formulas[["A"]], diamonds)
summary(mod)$sigma
mod$rank #i.e. degrees of freedom  = # vectors in colsp[X] to project onto
#Model B
mod = lm(model_formulas[["B"]], diamonds)
summary(mod)$sigma
mod$rank #i.e. degrees of freedom  = # vectors in colsp[X] to project onto
#Model C
mod = lm(model_formulas[["C"]], diamonds)
summary(mod)$sigma
mod$rank #i.e. degrees of freedom  = # vectors in colsp[X] to project onto
#Model D
mod = lm(model_formulas[["D"]], diamonds)
summary(mod)$sigma
mod$rank #i.e. degrees of freedom  = # vectors in colsp[X] to project onto
#Model E 
mod = lm(model_formulas[["E"]], diamonds)
summary(mod)$sigma
mod$rank #i.e. degrees of freedom  = # vectors in colsp[X] to project onto
```

Big win on E... the reason I think is that now we have a flexible curve for each continuous feature and that curve changes with all the categorical features.

Create model (F) which is the same as before except also include also third degree polynomials of the continuous features interacted with the categorical features and gauge performance against (E). By this time you're getting good with R's formula syntax!


```{r}
#Model F
model_formulas[["F"]] = price ~ 
        (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table + poly(carat, 3) + poly(x, 3) + poly(y, 3) + poly(z, 3) + poly(depth, 3) + poly(table, 3)) * (cut + color + clarity)
mod = lm(model_formulas[["F"]], diamonds)
summary(mod)$sigma
mod$rank
```

Even better (seemingly).

Can you think of any other way to expand the candidate set curlyH? Discuss.

We now have a very flexible curve for each continuous feature and that curve is fit individually for each level of all categories. We could indeed go further by allowing the levels among cut, color and clarity to interact and further allowing the continous features to interact. 

There is another strategy to increase the complexity of H without using function transforms of the features and interactions... we will have to see after winter break...

We should probably assess oos performance now. Sample 4,000 diamonds and use these to create a training set of 3,600 random diamonds and a test set of 400 random diamonds. Define K and do this splitting:

```{r}
K = 10
n_sub = 4000
set.seed(1984)
n_test = 1 / K * n_sub
n_train = n_sub - n_test
test_indicies = sample(1 : n, n_test)
train_indicies = sample(setdiff(1 : n, test_indicies), n_train)
all_other_indicies = setdiff(1 : n, c(test_indicies, train_indicies))
```

Compute in and out of sample performance for models A-F. Use s_e as the metric (standard error of the residuals). Create a list with keys A, B, ..., F to store these metrics. Remember the performances here will be worse than before since before you're using nearly 52,000 diamonds to build a model and now it's only 3600.

```{r}
oos_se = list()
all_models_train = list()
for (model_idx in LETTERS[1 : 6]){
  all_models_train[[model_idx]] = lm(model_formulas[[model_idx]], diamonds[train_indicies, ])
  summary(all_models_train[[model_idx]])$sigma
  oos_se[[model_idx]] = sd(diamonds$price[test_indicies] - predict(all_models_train[[model_idx]], diamonds[test_indicies, ]))
}
oos_se
```

You computed oos metrics only on n_* = 400 diamonds. What problem(s) do you expect in these oos metrics?

They are variable. And something is wrong with F! Possibly Runge's phenomenon?

Let's get a more stable estimate via K-fold cross validation:

```{r}
set.seed(1984)
temp = rnorm(n_sub)
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
rm(temp)
head(folds_vec, 100)
```

Do the K-fold cross validation for all models and compute the overall s_e and s_s_e. 

```{r}
oos_se = list()
oos_s_se = list()
for (model_idx in LETTERS[1 : 6]){
  e_vec_k = list() #for each one
  for (k in 1 : K){
    test_indicies_k = which(folds_vec == k)
    train_indicies_k = which(folds_vec != k)
    mod = lm(model_formulas[[model_idx]], diamonds[train_indicies_k, ])
    e_vec_k[[k]] = sd(diamonds$price[test_indicies_k] - predict(mod, diamonds[test_indicies_k, ]))
  }
  oos_se[[model_idx]] = mean(unlist(e_vec_k)) #note: not exactly the overall sd, but close enough
  oos_s_se[[model_idx]] = sd(unlist(e_vec_k))
}
res = rbind(unlist(oos_se), unlist(oos_s_se))
rownames(res) = c("avg", "sd")
res
```

Model F seems more variable than all of them. Why? Possibly Runge's phenomenon?

Imagine using the entire rest of the dataset besides the 4,000 training observations divvied up into slices of 400. Measure the oos error for each slice and also plot it.

```{r}
n_step = 1 / K * n_sub
oos_se = list()
ses = list()
starting_ks = seq(from = 1, to = (length(all_other_indicies) - n_step), by = n_step)
for (model_idx in LETTERS[1 : 6]){
  se_k = list() #for each one
  for (k in 1 : length(starting_ks)){
    diamonds_k = diamonds[all_other_indicies[starting_ks[k] : (starting_ks[k] + n_step - 1)], ]
    se_k[[k]] = sd(diamonds_k$price - predict(all_models_train[[model_idx]], diamonds_k))
  }
  ses[[model_idx]] = se_k
  oos_se[[model_idx]] = unlist(se_k)
}

pacman::p_load(reshape2)
ggplot(reshape2::melt(oos_se)) + geom_boxplot(aes(x = L1, y = value)) + xlab("model")
ggplot(reshape2::melt(oos_se)) + geom_boxplot(aes(x = L1, y = value)) + xlab("model") + ylim(0, 5000)
```

What do we learn from this? Model F is very risky. What exactly is going wrong? Part of the agony of data science is the "debugging" just like when debugging software. But here the errors can be in the modeling too! 

```{r}
max(unlist(ses[["F"]]))
k_BAD = which.max(unlist(ses[["F"]]))

diamonds_BAD = diamonds[all_other_indicies[starting_ks[k_BAD] : (starting_ks[k_BAD] + n_step - 1)], ]
diamonds_BAD

e_BAD = diamonds_BAD$price - predict(all_models_train[[model_idx]], diamonds_BAD)
tail(sort(abs(e_BAD)))
diamonds_BAD[which.max(abs(e_BAD)), ]

summary(diamonds[train_indicies, ])
```

Is it extrapolation? Yes... y = 58.9 and in a cubic function, that would be astronomical. The real mistake is that y = 58.9 is impossible. The data wasn't cleaned! 

**** Cleaning data essentially means visualizing each feature and the response (histograms for numeric and barplots for categorical) and ensuring there's no mistakes.

But this happens all the time and you don't want to be using polynomial functions for this reason since the new data may extrapolate very poorly.



## Model Selection and Three Data Splits

This unit is split into three use cases (1) Selection among M explicit models (2) Hyperparameter Selection within one algorithm (3) Stepwise Model Construction

# Use Case (I) Selecting one of M Explicit Models

We have now covered non-linearities (e.g. polynomial terms) and interactions. A new complication now clearly emerges. If I have $p$ predictors, there are many linear least squares models I can build (considering non-linear least squares models makes the space of models even larger!!)

For instance, here are a bunch of models and their in-sample performance:

```{r}
pacman::p_load(ggplot2)
all_model_formulas = list( #note: need these as strings for later...
  "price ~ carat + depth",
  "price ~ carat + depth + color + x + y + z",
  "price ~ .",
  "price ~ . * ."
)
mods = lapply(all_model_formulas, lm, diamonds)
lapply(mods, function(mod){summary(mod)$sigma})
```

Obviously the in-sample RMSE's are increasing due to the complexity, but which model is "best"?

This is one of the most fundamental problems in statistics, and possibly all of science! 

In class, we discussed validation via dividing $\mathbb{D}$ into (a) a training set and a (b) testing set. Now, we will further divide the training set into (a) a sub-training set and a (b) selection set and we still have the (c) test set. 

The total training set together will fit a model and testing will estimate future performance. But within the total training set, we'll use an elaborate algorithim: we'll fit many models and take the best one. That's the "master algorithm".

We'll make the selection set and the test set the same size but we don't have to. First split up the data:

```{r}
n = nrow(diamonds)
K = 5
set.seed(1984)
test_indices = sample(1 : n, size = n * 1 / K)
master_train_indices = setdiff(1 : n, test_indices)
select_indices = sample(master_train_indices, size = n * 1 / K)
train_indices = setdiff(master_train_indices, select_indices)
rm(master_train_indices)

#make sure we did this right:
pacman::p_load(testthat)
expect_equal(1 : n, sort(c(train_indices, select_indices, test_indices)))

diamonds_train = diamonds[train_indices, ]
diamonds_select = diamonds[select_indices, ]
diamonds_test = diamonds[test_indices, ]

rm(test_indices, select_indices, train_indices)
```

Now, fit all models and select the best one:

```{r}
mods = lapply(all_model_formulas, lm, diamonds_train)
```

Now predict on the selection set and look at the oos $s_e$, and select the "best" model

```{r}
yhat_selects = lapply(mods, function(mod){predict(mod, diamonds_select)})
y_select = diamonds_select$price #the true prices

s_e_s = lapply(yhat_selects, function(yhat_select){sd(yhat_select - y_select)})
s_e_s
#find the minimum
which.min(s_e_s)
```

Which are overfit? Which are underfit? Were these models "poor choices"?

Can we go back and fit some more models? 

Yes - as long as we don't open the "lockbox" of the test set. Let's look at one more model. An expansion of the best of the previous 4 models now with a couple interactions we are convinced are real plus a couple of non-linear terms:

```{r}
all_model_formulas[[5]] = "price ~ . + carat * color + carat * depth + I(carat^2) + I(depth^2)"
mods[[5]] = lm(all_model_formulas[[5]], diamonds_train) 

yhat_selects[[5]] = predict(mods[[5]], diamonds_select)

s_e_s[[5]] = sd(yhat_selects[[5]] - y_select)
s_e_s
#find the minimum
which.min(s_e_s)
```

We can go further and fit more and more models but we should always be careful that we don't fit too many as we may optimize to the selection set. Here, we are lucky since the selection set is large (~11,000 observations) so this is not too much of a fear.

But you can see the problem - how can we build a good model??

The answer to this is non-parametric regression / machine learning. But first, we will cover two other important topics before we get there.

Let us return and complete the exercise by now declaring we are done modeling and we are going to ship model 5. Let us get a conservative estimate of its performance:

```{r}
mod5_for_test = lm(all_model_formulas[[5]], rbind(diamonds_train, diamonds_select))
yhat_test_mod5 = predict(mod5_for_test, diamonds_test)
y_test = diamonds_test$price #the true prices
sd(yhat_test_mod5 - y_test)
```

About the same as the selection estimate --- we did not overfit too much to the selection set.

At this point the lockbox is open and we can never return (if we are honest, of course - many people in this business are not but it's never worth it).

Now we can build production model 4 with all data to ship:

```{r}
mod_final = lm(all_model_formulas[[5]], diamonds)
```

No evaluation is done on the final model. It is only used to predict future diamonds' prices.

We can make two improvements using CV to the above model selection procedure:

* To reduce variance in the selection process, you make a CV of the selection set. 
* To reduce variance in the testing process, you make an outer CV of the test set so that the first CV is a nested resampling. 

This is a lot more coding! But we're in luck because it's done for us already. I will eventually demo the `mlr3` package which makes this easy because it's too difficult to code from scratch without bugs.


# Use Case (II) Hyperparameter Selection

Remember the `lambda` from the Vapnik Objective Function which we minimized to compute an SVM for non-linearly separable data? We now have a way to automatically select its value. Each lambda value implies a different model. Hence we need to do "model selection" to select a hyperparameter which selects our model with the best performance.

Let's demo this on the `adult` dataset. Due to computational concerns only, let's limit the dataset to be size n = 2000 with 1/3-1/3-1/3 training-select-test sets.

```{r}
pacman::p_load_gh("coatless/ucidata") #load from github
data(adult)
adult = na.omit(adult)
adult$native_country = NULL
adult$income = factor(adult$income)
n = 2000
set.seed(5)
adult = adult[sample(1 : nrow(adult), n), ]
adult_train =  adult[1 : (n / 3), ]
adult_select = adult[(n / 3 + 1) : (2 * n / 3), ]
adult_test =   adult[(2 * n / 3 + 1) : n, ]
```

We now load the SVM package. Recall how to use the SVM: you specify a "cost" argument. That cost is related to the lambda value.

```{r}
pacman::p_load(e1071)
#e.g. 
#svm(income ~ ., adult_train, kernel = "linear", cost = 0.1)
```

So now we need to pick a bunch of cost values to search over. This is called a "grid search". It's still a decision! How many values in the grid? Step size? Minimum? Maximum? 

Let's search over M models. And we can specify cost values on the log10 scale.

```{r}
M = 30
cost_grid = 10^seq(from = -3, to = 4, length.out = M)
cost_grid
```

Let's do our search and collect errors on the select set:

```{r}
select_set_misclassification_errors_by_m = array(NA, M)
for (m in 1 : M){
  cat(m, "of", M, "\n")
  #train on train set
  svm_mod = svm(factor(income) ~ ., adult_train, kernel = "linear", cost = cost_grid[m])
  #predict on select set
  y_hat = predict(svm_mod, adult_select)
  #measure error from the select set
  select_set_misclassification_errors_by_m[m] = mean(adult_select$income != y_hat)
}
```

Let's plot these oos misclassification error rates:

```{r}
ggplot(data.frame(cost = cost_grid, miscl_err = select_set_misclassification_errors_by_m)) + 
  aes(x = cost, y = miscl_err) +
  geom_line(color = "grey") + 
  geom_point(lwd = 3) + 
  scale_x_log10()
```

Looks like the grid gave us a natural minimum. We can find the best value of the cost hyperparameter and its associated misclassification error:

```{r}
min(select_set_misclassification_errors_by_m)
optimal_cost_hyperparam = cost_grid[which.min(select_set_misclassification_errors_by_m)]
optimal_cost_hyperparam
```

Assume we will use this model and not test any more values of cost, we can get an honest performance metric for this optimally-tuned hyperparameter svm for the future on the test set:

```{r}
svm_mod = svm(income ~ ., adult_train, kernel = "linear", cost = optimal_cost_hyperparam)
y_hat = predict(svm_mod, adult_test)
mean(adult_test$income != y_hat)
```

This is about the same as we found on the select set. Thus, we didn't overfit the select set too much.

To ship the final model, use all the data.

```{r}
g_final = svm(income ~ ., adult, kernel = "linear", cost = optimal_cost_hyperparam)
```

We do not test the final model; we only use it to predict in the future.

# Use Case (III) Forward Stepwise Model Construction

There are many types of such stepwise models. Here we will look at Forward Stepwise Linear models. "Forward" meaning we start with a low complexity model and end with a high complexity model, "Stepwise" meaning we do so iteratively which each step consisting of one additional degree of freedom i.e. one incremental increase in complexity and "Linear" meaning that the model is linear. By default we use OLS.

We will be using the diamonds data again as an example. Let's make sure we have unordered factors to avoid issues later:

```{r}
rm(list = ls())
diamonds = ggplot2::diamonds
diamonds$cut = factor(diamonds$cut, ordered = FALSE)
diamonds$color = factor(diamonds$color, ordered = FALSE)
diamonds$clarity = factor(diamonds$clarity, ordered = FALSE)
```

What we're doing will be highly computational, so let's take a random sample of the diamonds in $\mathbb{D}$ for training, selecting and testing:

```{r}
Nsamp = 1300
set.seed(1984)
subindices = sample(1 : nrow(diamonds), Nsamp * 3)
diamonds_train = diamonds[subindices[1 : Nsamp], ]
diamonds_select = diamonds[subindices[(Nsamp + 1) : (2 * Nsamp)], ]
diamonds_test = diamonds[subindices[(2 * Nsamp + 1) : (3 * Nsamp)], ]
rm(subindices)
```

Let's built a model with all second-order interactions e.g. all things that look like depth x table x clarity or depth^2 x color or depth^3.

```{r}
mod = lm(price ~ . * . * ., diamonds_train)
```

How many variables is this? And what does it look like?

```{r}
length(coef(mod))
```

Here's what some of the feature names look like:

```{r}
sample(names(coef(mod)), 100)
```

For features that are non-binary, the total will be p_non_binary^3 features. Binary features are more complicated because its each level in feature A times each level in feature B. There are no squared or cube terms for binary features (since they're all the same i.e. ${0,1}^d = {0,1}$).

Remember we likely overfit just using first order interactions? We'll certainly overfit using first-order interactions AND second order interactions.

```{r}
summary(mod)$r.squared
sd(summary(mod)$residuals)
```

Is that believable? Well... let's try predicting using this insano model on another Nsamp we didn't see...

```{r}
y_hat_test = predict(mod, diamonds_test)
y_test = diamonds_test$price
e_test = y_test - y_hat_test
1 - sum((e_test)^2) / sum((y_test - mean(y_test))^2)
sd(e_test)
```

VERY negative oos $R^2$ and awful RMSE --- why? We overfit big time!

```{r}
sd(y_test)
sd(e_test) / sd(y_test)
```

This is not only "overfitting"; it is an absolute trainwreck! This means you can do ~60x better using the null model (average of y) instead of this model.

So let us employ stepwise to get a "good" model. We need our basis predictors to start with. How about the linear components of `. * . * .` --- there's nothing intrinsically wrong with that - it's probably a good basis for $f(x)$. Let's create the model matrix for both train and test:

```{r}
Xmm_train = model.matrix(price ~ . * . * ., diamonds_train)
y_train = diamonds_train$price
p_plus_one = ncol(Xmm_train)
Xmm_select = model.matrix(price ~ . * . * ., diamonds_select)
y_select = diamonds_select$price
```

Now let's go through one by one and add the best one based on $s_e$ gain i.e. the best new dimension to add to project the most of the vector $y$ as possible onto the column space.

```{r}
included_features_by_iter = c() #keep a growing list of predictors by iteration
in_sample_ses_by_iteration = c() #keep a growing list of se's by iteration
oos_ses_by_iteration = c() #keep a growing list of se's by iteration
i = 1

repeat {

  #get all predictors left to try
  all_ses = array(NA, p_plus_one) #record all possibilities
  for (j_try in 1 : p_plus_one){
    if (j_try %in% included_features_by_iter){
      next 
    }
    Xmm_sub = Xmm_train[, c(included_features_by_iter, j_try), drop = FALSE]
    all_ses[j_try] = sd(lm.fit(Xmm_sub, y_train)$residuals) #lm.fit so much faster than lm! 
  }
  j_star = which.min(all_ses)
  included_features_by_iter = c(included_features_by_iter, j_star)
  in_sample_ses_by_iteration = c(in_sample_ses_by_iteration, all_ses[j_star])
  
  #now let's look at oos
  Xmm_sub = Xmm_train[, included_features_by_iter, drop = FALSE]
  mod = lm.fit(Xmm_sub, y_train)
  y_hat_select = Xmm_select[, included_features_by_iter, drop = FALSE] %*% mod$coefficients
  oos_se = sd(y_select - y_hat_select)
  oos_ses_by_iteration = c(oos_ses_by_iteration, oos_se)
  
  cat("i =", i, "in sample: se = ", round(all_ses[j_star], 1), "oos_se", round(oos_se, 1), "added:", colnames(Xmm_train)[j_star], "\n")
  
  i = i + 1
  
  if (i > Nsamp || i > p_plus_one){
    break #why??
  }
}
```

Now let's look at our complexity curve:

```{r}
simulation_results = data.frame(
  iteration = 1 : length(in_sample_ses_by_iteration),
  in_sample_ses_by_iteration = in_sample_ses_by_iteration,
  oos_ses_by_iteration = oos_ses_by_iteration
)

pacman::p_load(latex2exp)
ggplot(simulation_results) + 
  geom_line(aes(x = iteration, y = in_sample_ses_by_iteration), col = "red") +
  geom_line(aes(x = iteration, y = oos_ses_by_iteration), col = "blue") + 
  ylim(0, max(c(simulation_results$in_sample_ses_by_iteration, simulation_results$oos_ses_by_iteration)))
  ylab(TeX("$s_e$"))
```


We can kind of see what the optimal model is above. There is a whole zone of flat models. Better to be simple. So let's pick it from the graph visually. You can also obviously do CV within each iterations to stabilize this further.

What's the optimal number of features? And its associated oos s_e?

```{r}
p_opt = 41 #change this based on plot above
oos_ses_by_iteration[p_opt]
```

What are those features?

```{r}
sort(colnames(Xmm_train)[included_features_by_iter[1 : p_opt]])
```

Lots of carat-color-clarity interactions which makes sense. Other features don't matter so much as their information is probably duplicated within carat, color and clarity.

What is the "true optimal model"? It's impossible to find as we'd have to search all subsets (of which there are exponential number of them 2^1000 = 10^31). Here we used a greedy search so hopefully we get somewhere in the ballpark of optimal. 

We now get a conservative estimate of its future performance:

```{r}
Xmm_test = model.matrix(price ~ . * . * ., diamonds_test)
optimal_mod = lm(diamonds_test$price ~ 0 + Xmm_test[, included_features_by_iter[1 : p_opt]])
summary(optimal_mod)$sigma
```

Looks even better than we found during the stepwise iterations on the select set. This is random variation. It would be better to use CV within the stepwise and CV outside to then stabilize this final estimate.

The final step is to fit g_final on all data using those selected features.




## Piping

Take a look at this one-liner:


```{r}
set.seed(1984)
mean(head(round(sample(rnorm(1000), 100), digits = 2)))
```

This is hard to read. Of course we can make it easier by using breaklines e.g.

```{r}
mean(
  head(
    round(
      sample(
        rnorm(1000), 
        100
      ), 
      digits = 2
    )
  )
)
```

But it doesn't make it much easier to read. And it probably makes it harder to write. 

Enter an idea taken from unix / linux. Output of one function is input to next function. It is the inverse of the usual "order of operations". Let's see how this works.

We first load the piping library:

```{r}
pacman::p_load(magrittr)
```

The package is named after Rene Magritte, the Belgian surrealist artist because he wrote [(Ceci n'est pas un pipe)](https://en.wikipedia.org/wiki/The_Treachery_of_Images) on a painting of a pipe.

In pipe format this would look like:

```{r}
set.seed(1984)
rnorm(1000) %>% #the pipe operator
  sample(100) %>% 
  round(digits = 2) %>% #the first argument is passed in automatically.
  head %>%
  mean
```

That's it! There's nothing more to it other than a gain in readability. Here's a cute joke based on this idea:

https://www.mzes.uni-mannheim.de/socialsciencedatalab/article/efficient-data-r/figures/basertidyverse.png

What if we wanted to do something like `mean(rnorm(1000) + 1)`? This `rnorm(1000) %>% +1 %>% mean` doesn't work because I imagine because the basic arithmetic operators couldn't be parsed like normal while there was a pipe. So they invented special pipe functions for this:

```{r}
rnorm(1000) %>% 
  add(1) %>% 
  mean
```

There are other exceptions to the rule too which you'll figure out if you adopt the pipe.

Unfortunately... the world at large hasn't completely accepted this as a way to write R. So feel free to use for yourself. But be careful when using this style with others. There are places where everyone uses the pipes (we will see this when we get to dplyr). Also note the code you write with pipes will be slower than the normal syntax.

# The Grammar of graphics and ggplot

First load the package and the dataset of interest as a dataframe:

```{r}
pacman::p_load(ggplot2, quantreg)
cars = MASS::Cars93 #dataframe
```

ggplot is based on the "Grammar of Graphics", a concept invented by the Statistician / Computer Scientist Leland Wilkinson who worked on SPSS, Tableau and now he works at H20, software that analyzes big data. The reference of interest is [here](http://papers.rgrossman.com/proc-094.pdf). He drew on ideas from John Tukey (one of the great statistician of the previous generation) while he was at Bell Labs, Andreas Buja (one of my professors at Penn) and Jerome Friedman (the professor that taught my data mining course when I was in college at Stanford). 

It is a language that allows us to describe the components of a graphic. Previously, graphics were done in one shot and it was clunky. ggplot is a library written by Hadley Wickham based on this concept. Wickham is probably the most famous person in statistical computing today. He has commit rights in R and is one of the architects of RStudio. He calls grammar of graphics "graphical poems". Here are the basic components:

* an underlying data frame
* an "aesthetic" that maps visualization axes in the eventual plot(s) to variables in the data frame
* a "layer" which is composed of
  - a geometric object
  - a statistical transformation
  - a position adjustment
* a "scale" for each aesthetic
* a "coordinate" system for each aesthetic
* optional "facets" (more graphics)
* optional "labels" for the title, axes title, points, etc.

Don't worry - everything has "smart defaults" in Wickham's implementation so you don't have to worry about most things. We will explore some of the features below. Here's a good [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

ggplot is layered where each component is an object. The objects are added to each other since the "+" operator is overloaded to accept these additions. This is nice because each component can be saved and reused. The following initialized the graphics:

```{r}
ggplot(cars)
```

Nothing happened - except the plot window went gray (the smart default). This is already rendering a graphic, but since it hasn't been given any of the required information, it has nothing to display. Next we create an aesthetics indicating a one-way plot (one variable only).

```{r}
ggplot(cars) + 
  aes(Price)
```

Notice how it can understand the variable name as an object name.

Since we've given it an aesthetics object, it now knows which variable is the x axis (default). It already knows the ranges of the variable (a smart default) and a default scale and coordinate system (smart defaults).

Usually this is done in one step by passing the aesthetics object into the ggplot:

```{r}
ggplot(cars, aes(Price))
```

Now we need to pick a layer by specifying a geometry. This is a type of plot. Since the predictor type of price is continuous, let's pick the "histogram" using the `geom_histogram` function:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram()
```

This can be customized:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(binwidth = 1, col = "darkred", fill = "blue", alpha = 0.4)
```

Want to save it for your latex?

```{r}
ggsave("plot.png")
system("open plot.png")
ggsave("plot.pdf")
system("open plot.pdf")
```

Here are some other options besides the histogram:

```{r}
ggplot(cars, aes(Price)) +
  geom_dotplot()
ggplot(cars, aes(Price)) +
  geom_area(stat = "bin", binwidth = 2)
ggplot(cars, aes(Price)) +
  geom_freqpoly()
ggplot(cars, aes(Price)) +
  geom_density(fill = "green", alpha = 0.1)

summary(cars)
```


Can we compare price based on different conditions? Yes, we can subset the data and use color and alpha:

```{r}
ggplot(cars, aes(Price)) +
  geom_density(data = subset(cars, Man.trans.avail == "Yes"), col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_density(data = subset(cars, Man.trans.avail == "No"), col = "grey", fill = "red", alpha = 0.4)
```

Sidebar: why are cars that have manual transmissions available cheaper?

We can look at this also using a histogram of the conditional distributions:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(data = subset(cars, Man.trans.avail == "Yes"), binwidth = 1, col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_histogram(data = subset(cars, Man.trans.avail == "No"), binwidth = 1, col = "grey", fill = "red", alpha = 0.4)
```

What if the variable is not continuous e.g. Cylinders? We can use a bar graph / bar plot.

```{r}
ggplot(cars, aes(Cylinders)) +
  geom_bar()
```

This is essential frequency by level of the categorical variable.

Now let's move on to looking at one variable versus another variable. For example price by engine power:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price))
```

Since we've given it an aesthetics object, it now knows which variable is the x axis and which variable is the y axis. It already knows the ranges of the variables (a smart default) and a default scale and coordinate system (smart defaults).

Just as before, now we need to pick a layer by specifying a geometry. This is a type of plot. Let's pick the "scatterplot" using the `geom_point` function:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point()
```

Now we have a nice scatterplot. This function uses the inherited data, the inherited aesthetics. Since this "geometry" is a "layer", we can pass in options to the layer.

```{r}
base_and_aesthetics = ggplot(cars, aes(x = Horsepower, y = Price))
base_and_aesthetics + 
  geom_point(col = "red", fill = "green", shape = 23, size = 3, alpha = 0.3)
```

Let's handle names of axes, title and ranges:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics +
  ggtitle("Average Car Price vs. Engine Power", subtitle = "in the Cars93 dataset") +
  ylab("Price (in $1000's)")
base_and_aesthetics_with_titles +
  geom_point() +
  xlim(0, 400) +
  ylim(0, 50)
  
```

Let's transform the variables:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_x_continuous(trans = "log2")
```

Each unit increase on the x-axis now represent a doubling increase in x (although the whole scale only spans 3 units). But look at how the grid didn't keep up. Let's fix this:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_x_continuous(trans = "log2", breaks = round(seq(0, max(cars$Horsepower), length.out = 6)))
```

We can do the same to the y axis:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_y_continuous(trans = "log10")+
  scale_x_continuous(trans = "log10")
```

Let's look at some more geometries.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth()
```

Here, I've added two geometries on the same aesthetic! This attempts to explain the relationship $f(x)$ using smoothing. Let's go for more.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_rug()
```

This allows us to also see the marginal distributions of the two variables.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
```

This fits a line and tries to indicate statistical significance of the line. We have *not* covered any statistics in this class yet (ironic!) ... so ignore how the window is generated.

Can we display more than two dimensions? Yes. Let's indicate a third dimension with shape (only works with factors).

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power and Transmission")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail)) +
  geom_smooth() +
  geom_rug()
```

Can we display more than three dimensions? Yes. Let's indicate a fourth dimension with color.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain)) +
  geom_smooth() +
  geom_rug()
```

Can we go to a fifth dimension? Maybe?

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, size = Weight), alpha = 0.5) + #size?
  geom_smooth() +
  geom_rug()
```

A seventh? We can use text labels adjacent to the scatterplot's points.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission, Drivetrain,  Weight & #Cylinders")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, alpha = Weight)) + #size?
  geom_text(aes(label = Cylinders), vjust = 1.5, col = "darkgrey", lineheight = 0.3, size = 3) +
  geom_smooth() +
  geom_rug()
```

Getting difficult to see what's going on.

Let's move away from the scatterplot to just density estimation:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power") #reset the title
base_and_aesthetics_with_titles +
  geom_density2d()
```

Other alternatives:

```{r}
base_and_aesthetics_with_titles +
  geom_bin2d(binwidth = c(8, 3))
pacman::p_load(hexbin)
base_and_aesthetics_with_titles +
  geom_hex()
```

This is like a two-dimensional histogram where the bar / hexagon heights are seen with color.

What if the x-axis is categorical for example Cylinders versus price? Typical is the "box and whiskers" plot:

```{r}
ggplot(cars, aes(x = Cylinders, y = Price)) +
  geom_boxplot()
```

Clear relationship!

How about multiple subplots based on the subsetting we did in the histograms? This is called "faceting". Here are two bivariate visualizations laid horizontally:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(. ~ Man.trans.avail)
```

Or alternatively, vertically:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(Man.trans.avail ~ .)
```

And we can even double-subset:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin)
```

And we can even triple-subset or more:

```{r}
cars$MedWeight = ifelse(cars$Weight > median(cars$Weight), ">MedWeight", "<MedWeight")
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin + MedWeight, scales = "free")
```

These three varibles seem somewhat independent.

There are other primitives like `geom_abline` which graphs a line and `geom_segment` we will see today. Note that if you want plots rendered within functions or loops you have to explicitly call the `plot` function:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + 
    geom_histogram(aes(x))
  graphics_obj
}
```

versus:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + geom_histogram(aes(x))
  plot(graphics_obj)
}
```


Lastly, ggplot offers lots of nice customization themes:

```{r}
graphics_obj = base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
graphics_obj + theme_bw()
graphics_obj + theme_dark()
graphics_obj + theme_classic()

```

Packages offer even more:

```{r}
pacman::p_load(forcats, lazyeval, ggthemes)
graphics_obj + theme_economist()
graphics_obj + theme_stata()
graphics_obj + theme_tufte()
```

and of course, the whimsical one and only:


```{r}
pacman::p_load(xkcd, extrafont)
download.file("http://simonsoftware.se/other/xkcd.ttf", dest = "xkcd.ttf", mode = "wb")
#MAC
# system("mv xkcd.ttf /Library/Fonts")
# font_import(path = "/Library/Fonts", pattern = "xkcd", prompt = FALSE)
# fonts()
# fonttable()
# loadfonts()
#WINDOWS
font_import(path = ".", pattern = "xkcd", prompt = FALSE)
fonts()
fonttable()

loadfonts(device="win")

graphics_obj + theme_xkcd()
```



# C++ and R

R goes back to 1995 when it was adapted from S (written in 1976 by John Chambers at Bell Labs) with minor modifications. The core of base R is written in C and Fortran. These two languages are the fastest known languages (how to measure "fastest" is a huge debate). Thus, base R is very fast. For instance the `sort` function is as fast as C/Fortran since it immediately calls compiled C/Fortran routines.

However, R code itself that you write is "interpreted" which means it is not compiled until you run it. And it has to compile on-the-fly, making it very slow. Prior to v3.4 (April, 2017) it was even slower since the code wasn't JIT compiled. All this "real CS" stuff you can learn in another class..

One notable place to observe this slowness relative to other languages is in looping. For example:

```{r}
SIZE = 1e6
v = 1 : SIZE
```

Take for example a simple function that computes square roots on each element

```{r}
sqrt_vector = function(v){
  v_new = array(NA, length(v))
  for (i in 1 : length(v)){
    v_new[i] = sqrt(v[i])
  }
  v_new
}
```

How fast does this run? Let's use a cool package called `microbenchmark` that allows us to do an operation many times and see how long it takes each time to get an average:

```{r}
pacman::p_load(microbenchmark)
microbenchmark(
  sqrt_vector(v), 
  times = 10
)
```

Does the apply function help?

```{r}
microbenchmark(
  apply(v, 1, FUN = sqrt), 
  times = 10
)
```

Strange that this takes so long? So it doesn't help... it hurts A LOT. Unsure why... Be careful with apply! 

How much faster in C++ should this be?

Enter the `Rcpp` package - a way to compile little bits (or lotta bits) of C++ on the fly.

```{r}
pacman::p_load(Rcpp)
```


Let's write this for loop function to sqrt-ize in C++. We then  compile it and then save it into our namespace to be called like a regular function. Note that we use C++ classes that are not part of standard C++ e.g. "NumericVector". Rcpp comes build in with classes that are interoperable with R. It's not hard to learn, just takes a small dive into the documentation.

```{r}
cppFunction('
  NumericVector sqrt_vector_cpp(NumericVector v) {
    int n = v.size();
    NumericVector v_new(n);
    for (int i = 0; i < n; i++) { //indices from 0...n-1 not 1...n!
      v_new[i] = sqrt(v[i]);
    }
    return v_new;
  }
')
```

What do these two functions look like?

```{r}
sqrt_vector
sqrt_vector_cpp
```

The first one shows the R code and then says it is bytecode-compiled which means there are speedups used in R (go to an advanced CS class) but we will see these speedups aren't so speedy! The other just says we `.Call` some C++ function in a certain address (pointer) and the argument to be inputted.

What is the gain in runtime?

```{r}
microbenchmark(
  sqrt_vector_cpp(v), 
  times = 10
)
```

WOW. 10x!!! Can't beat that with a stick...

Let's do a not-so-contrived example...

Matrix distance... Let's compute the distances of all pairs of rows in a dataset. I will try to code the R as efficiently as possible by using vector subtraction so there is only two for loops. The C++ function will have an additional loop to iterate over the features in the observations.

```{r}
#a subset of the diamonds data
SIZE = 1000
X_diamonds = as.matrix(ggplot2::diamonds[1 : SIZE, c("carat", "depth", "table", "x", "y", "z")])

compute_distance_matrix = function(X){
  n = nrow(X)
  D = matrix(NA, n, n)
  for (i_1 in 1 : (n - 1)){
    for (i_2 in (i_1 + 1) : n){
      D[i_1, i_2] = sqrt(sum((X[i_1, ] - X[i_2, ])^2))
    }
  }
  D
}

cppFunction('
  NumericMatrix compute_distance_matrix_cpp(NumericMatrix X) {
    int n = X.nrow();
    int p = X.ncol();
    NumericMatrix D(n, n);
    std::fill(D.begin(), D.end(), NA_REAL);

    for (int i_1 = 0; i_1 < (n - 1); i_1++){
      //Rcout << "computing for row #: " << (i_1 + 1) << "\\n";
      for (int i_2 = i_1 + 1; i_2 < n; i_2++){
        double sqd_diff = 0;
        for (int j = 0; j < p; j++){
          sqd_diff += pow(X(i_1, j) - X(i_2, j), 2); //by default the cmath library in std is loaded
        }
        D(i_1, i_2) = sqrt(sqd_diff); //by default the cmath library in std is loaded
      }
    }
    return D;
  }
')
```

```{r}
microbenchmark(
  {D = compute_distance_matrix(X_diamonds)},
  times = 10
)

round(D[1 : 5, 1 : 5], 2)
```

Slow...

```{r}
microbenchmark(
  {D = compute_distance_matrix_cpp(X_diamonds)},
  times = 10
)
round(D[1 : 5, 1 : 5], 2)
```

Absolutely lightning... ~200x faster on my laptop than R's runtime.

Writing functions as strings that compile is annoying. It is better to have separate files. For instance...

```{r}
sourceCpp("distance_matrix.cpp")
```

Here are a list of the data structures in Rcpp: https://teuder.github.io/rcpp4everyone_en/070_data_types.html#vector-and-matrix

Another place where C++ pays the rent is recursion. Here is a quicksort implementation in R taken from somewhere on the internet.

```{r}
quicksort_R <- function(arr) {
  # Pick a number at random.
  mid = sample(arr, 1)

  # Place-holders for left and right values.
  left = c()
  right = c()
  
  # Move all the smaller values to the left, bigger values to the right.
  lapply(arr[arr != mid], function(d) {
    if (d < mid) {
      left <<- c(left, d) #needs to assign to the global variable here to jump out of the scope of the apply function
    }
    else {
      right <<- c(right, d) #needs to assign to the global variable here to jump out of the scope of the apply function
    }
  })
  
  if (length(left) > 1) {
    left = quicksort_R(left)
  }
  
  if (length(right) > 1) {
    right = quicksort_R(right)
  }
  
  # Finally, return the sorted values.
  c(left, mid, right)
}
```

Let's create a random array to test these sorts on:

```{r}
n = 10000
x = rnorm(n)
```


Let's profile the pure R sort function:

```{r}
microbenchmark(
  x_sorted_pure_R = quicksort_R(x),
  times = 10
)
```

Let's profile R's `sort` function.

```{r}
microbenchmark(
  x_sorted_base_R = sort(x),
  times = 10
)
```

Let's just ensure our method worked...

```{r}
x_sorted_pure_R = quicksort_R(x)
x_sorted_base_R = sort(x)
pacman::p_load(testthat)
expect_equal(x_sorted_pure_R, x_sorted_base_R)
```

Basically infinitely faster. Let's make our own C++ implementation.

```{r}
sourceCpp("quicksort.cpp")
```

and profile it:

```{r}
microbenchmark(
  x_sorted_cpp = quicksort_cpp(x),
  times = 10
)
```

Let's just ensure this method worked...

```{r}
pacman::p_load(testthat)
expect_equal(x_sorted_cpp, x_sorted_base_R)
```

Why is our C++ slower than `sort`. Because `sort` is also in C++ or Fortran and it's been likely optimized and reoptimized up to wazoo for decades. Also, Rcpp's data structures may be slower than base R's data structures. There may be some speed lost to translating to `NumericVector` from `double[]` or something like that.

Can you call R from Rcpp? You bet:

```{r}
cppFunction('
  NumericVector rnorm_cpp_R(int n, double mean, double sd){
      // get a pointer to R\'s rnorm() function
      Function f("rnorm");   
  
      // Next code is interpreted as rnorm(n, mean, sd)
      return f(n, Named("sd")=sd, _["mean"]=mean);
  }
')

rnorm_cpp_R(5, 1, .01)
```

A few math functions are implemented for you already:

```{r}
evalCpp('R::qnorm(0.5, 0, 1, 1, 0)')
evalCpp('R::qnorm(0.5, 0, 1)') #BOOM
```

Further, there are many common functions that are already wrapped for you via "Rcpp-sugar" which was the Rcpp's author's attempt to make Rcpp a whole lot easier, see [here](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-sugar.pdf).

```{r}
evalCpp('rnorm(10, 100, 3)')
```

If you want blazing fast linear algebra, check out package `RcppArmadillo` which is a wrapper around Apache's Armadillo (namespace is "arma" in the code), an optimized linear algebra package in C++. Here is an example taken from [here](https://scholar.princeton.edu/sites/default/files/q-aps/files/slides_day4_am.pdf). It involves solving for b-vec in a standard OLS.

```{r}
pacman::p_load(RcppArmadillo)

cppFunction('
  arma::mat ols_cpp(arma::mat X, arma::mat y){
    arma::mat Xt = X.t();
    return solve(Xt * X, Xt * y);
  }
', depends = "RcppArmadillo")

n = 500
Xy = data.frame(int = rep(1, n), x1 = rnorm(n), x2 = rnorm(n), x3 = rnorm(n), y = rnorm(n))
X = as.matrix(Xy[, 1 : 4])
y = as.matrix(Xy[, 5])

#does the function work?
expect_equal(as.numeric(ols_cpp(X, y)), as.numeric(solve(t(X) %*% X) %*% t(X) %*% y))
```

Now how fast is it?

```{r}
microbenchmark(
  R_via_lm = lm(y ~ 0 + ., data = Xy),
  R_matrix_multiplication = solve(t(X) %*% X) %*% t(X) %*% y,
  cpp_with_armadillo = ols_cpp(X, y),
    times = 100
)
```

About 4x faster than R's optimized linear algebra routines. Supposedly it can go even faster if you enable parallelization within Armadillo. I couldn't get that demo to work...

Note lm is slow because it does all sorts of other stuff besides computing b-vec e.g. builds the model matrix, computes Rsq, computes residuals, does statistical testing, etc...

Here are the places where Rcpp is recommended to be used (from https://teuder.github.io/rcpp4everyone_en/010_Rcpp_merit.html)

* Loop operations in which later iterations depend on previous iterations.
* Accessing each element of a vector/matrix.
* Recurrent function calls within loops.
* Changing the size of vectors dynamically.
* Operations that need advanced data structures and algorithms (we don't do this in this class).

# Java and R

We just did C++ with R. Is there a bridge to Java? Yes (and there's bridges to many other languages too). Java and R can speak to each other through proper configuration of the `rJava` package. You need to have a full JDK of Java installed on your computer and have its binary executables in the proper path. This demo will be in Java JDK 8 (released in 2014 and not officially supported after 2020) since I haven't tested on the more modern Java JDK's yet. We first install `rJava` if necessary:

```{r}
if (!pacman::p_isinstalled(rJava)){
  pacman::p_load(pkgbuild)
  if (pkgbuild::check_build_tools()){
    install.packages("rJava", type = "source")
  }
  install.packages("rJava")
}
```

Now we load the package. Before we do, we set the JVM to have 8GB of RAM. After we load it, we initialize te JVM. This should print out nothing or "0" to indicate success.

```{r}
options(java.parameters = "-Xmx8g")
pacman::p_load(rJava)
.jinit() #this initializes the JVM in the background and if this runs with no issues nor output, you probably have rJava installed and connected to the JDK properly.
```

Just like the whole `Rcpp` demo, we can do a whole demo with `rJava`, but we won't. Here's just an example of creating a Java object and running a method on it:

```{r}
java_double = .jnew("java/lang/Double", 3.1415)
java_double
class(java_double)
.jclass(java_double)
#call an instance method 
.jcall(java_double, "I", "intValue") #java_double.intValue();
#call a static method
J("java/lang/String", "valueOf", java_double)
```

A note on rJava vs Rcpp. 

* If you're doing quick and dirty fast functions for loops and recursion, do it in Rcpp since there is lower overhead of programming. 
* If you are programming more full-featured software, go with rJava. 
* Also, if you need full-featured parallelized execution and threading control e.g. thread pooling and the ease of debugging, my personal opinion is that rJava is easier to get working with less dependencies. Rcpp threading is trickier and so is the openMP directives within Rcpp.
* Further, the JVM is fully asynchronous which means it runs completely independently of R. What this means is that you can execute something in Java, Java can "thread it off" and return you to the R prompt with a pointer to the object that houses its execution. You can then query the object. We will see demos of this.


# Python and R

No demo would be complete without this.

```{r}
pacman::p_load(reticulate)
py_available()
py_numpy_available()

# import numpy and specify no automatic Python to R conversion
np = import("numpy", convert = FALSE)

# do some array manipulations with NumPy
python_arr = np$array(1 : 4)
cumsum_python = python_arr$cumsum()
class(cumsum_python)
cumsum_python

# convert to R explicitly at the end
cumsum_R = py_to_r(cumsum_python)
cumsum_R

# now do the opposite, start with R and convert to Python
r_to_py(cumsum_R)
r_to_py(as.integer(cumsum_R))
```

Let's look at an example of Python Data Analysis Library (pandas). Let's install if not already installed:

```{r}
import("pandas", convert = FALSE)
# py_install("pandas")
```

And python even works in Rstudio's markdown files e.g.

```{python}
#this is python code!!!
import pandas as pd
flights = pd.read_csv("https://raw.githubusercontent.com/PacktPublishing/Pandas-Cookbook/master/data/flights.csv")
flights = flights.dropna()
flights.columns
flights[flights['DEST_AIR'] == "JFK"]
```

And then switch back to R and have access to the object we instantiated in python via the `py` object:

```{r}
#this is R code!!!
ggplot(py$flights) + 
  aes(x = AIRLINE, y = ARR_DELAY) +
  geom_boxplot()

lm(ARR_DELAY ~ AIRLINE, py$flights)
```

